The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-01-18 src/ms.tex
added governance section

Briefly summarize new procedures and governance structure, new CoCo, election
process overview, etc.

>>>>

%Briefly summarize new procedures and governance structure, new CoCo, election
%process overview, etc.

As part of the process of developing Astropy into a long-term sustainable product, and to improve transparency and accountability, the Project agreed to write down and formalize our governance structure (partly supported by explicit funding for this purpose - see \S \ref{sec:project-funding}). At the 2019 coordination meeting input was gathered from participants on what governance structures existed in the associated Open Source Software communities, and what would fit well with the needs of Astropy. This led into a ``retreat'' planned for March 2020, but due to the COVID-19 pandemic, this became a serious of virtual meetings of the ``Astropy Governance Working Group''.  This group drafted the APE0 document \citep{ape0}, which was then eventually ratified and implemented by the ``Astropy Governance Implementation Working Group'' in Fall 2021. While the process emphasized flexibility and the ability to adapt to changing circumstances, it is expected that this is the framework Astropy's governance will operate in for at least the medium-term future.

The APE0 \citep{ape0} document lays out the principles of this governance structure, so we refer the reader to that document for a more thorough description.  However here we highlight some key elements. While many of these principles were already de facto true or have been discussed organically (and have been discussed in earlier papers in this series), the APE0-based governance aimed to provide a single place where the community can agree as a starting point. With this in mind, it highlights the developer and user community of the Astropy Project as the ultimate sources of authority, as well as the core principle of ``do-ocracy'' that those who do work for the Project (be it coding, training, or other less concrete contributions) gain more influence on the outputs of the Project by virtue of their effort.  APE0 adds, however, the concept of ``voting members'' - a self-governed part of that community who are entrusted to elect the Coordination Committee. While this committee has existed from the inception of the project, APE0 establishes a formal voting process for this committee, and explicitly outlines the rights and responsibilities of the Coordination Committee. This role is mainly to facilitate consensus and act as the decision maker when other mechanisms have failed.  However, it also includes powers that either require central authority or secrets (e.g. passwords), but AOE0 also charges the Committee to devolve responsibilities and seek community input on these items as often as possible.

The first coordinating committee election under these rules took place in fall 2021, electing a mix of prior existing and new coordination committee members, and was contested in the sense of more candidates than available slots. This suggests the process is already working to serve the long-term interests of the committee to both spread the coordination effort, and to ensure it is not dominated by the same people for as long as the Project continues. While other, more fine-grained governance improvements are planned for the future, it is clear the foundation is now in place.
